---
title: "Prediction of recidivism in a long-term follow-up of forensic psychiatric patients: Incremental effects of neuroimaging data"
author: "Carl Delfin"
date: "10 january 2018"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Introduction

This is an RMarkdown document containing the R code used for all data analysis in the study *Prediction of recidivism in a long-term follow-up of forensic psychiatric patients: Incremental effects of neuroimaging data*. Externally sourced scripts are located in the *scripts* folder.

# Clear environment

Clear any existing data, values and functions from the global environment.

```{r}
rm(list = ls())
```

# Start timer

Not really necessary, but I want to keep track of total running time for the entire analysis pipeline.

```{r}
timer <- proc.time()
```

# Set-up

Initial set up, such as setting up paths, loading packages and reading data.

## Paths

My paths (e.g. to data) are stored in an external script for security reasons. Every variable written in ALL CAPS is a pre-defined path/key.

```{r}
source("secret/paths.R")
```

## Load packages 

All required packages are stored in a vector.

```{r}
requiredPackages <- c("foreign", "psych", "Barnard", "caret", "randomForest", "ROCR", "ggplot2", "cowplot", "pdp", "DescTools")
```

The `ipak` script goes through the *requiredPackages* list of packages and checks if they are installed, installs them if not, then loads them.

```{r}
source("scripts/ipak.R")
ipak(requiredPackages)
```

## Data and seed

The data is read from an SPSS file, and the seed is set for reproducibility (I always use the year the analysis was conducted, in this case 2017, to avoid "seed hacking").

```{r, message=FALSE, warning=FALSE}
data <- read.spss(DATALOCATION, to.data.frame = TRUE)
seed <- 2017
```

## Numeric to factor

Reading SPSS data sometimes turns factor variables to numeric ones, which must be manually corrected.

```{r}
# baseline risk factors
data$male <- as.factor(data$male)
data$SUD <- as.factor(data$SUD)
data$PDB <- as.factor(data$PDB)
data$education <- as.factor(data$education)
data$MDFDR <- as.factor(data$MDFDR)
data$prevcrimeall <- as.factor(data$prevcrimeall)

# primary DSM-IV diagnosis
data$primarydsmdiagnosis <- as.factor(data$primarydsmdiagnosis)
data$primarydsm_psychosis <- as.factor(data$primarydsm_psychosis)
data$primarydsm_mood <- as.factor(data$primarydsm_mood)
data$primarydsm_pd <- as.factor(data$primarydsm_pd)
data$primarydsm_DDA <- as.factor(data$primarydsm_DDA)
data$primarydsm_NDD <- as.factor(data$primarydsm_NDD)

# pharmacological treatment
data$antipsychotic <- as.factor(data$antipsychotic)
data$antidepressant <- as.factor(data$antidepressant)
data$benzodiazepine <- as.factor(data$benzodiazepine)
data$nonbenzodiazepine <- as.factor(data$nonbenzodiazepine)
data$anticholinergic <- as.factor(data$anticholinergic)

# other variables
data$stillundercare <- as.factor(data$stillundercare)

```

# Data pre-processing

## Select predictors

The outcome variable, predictor variables, and other variables of interest are defined and stored in vectors.

```{r}
# outcome variable
outcomeVariable <- "reoffendgeneral"

# baseline variables
baselineList <- c("age", "agefirstcrime", "PCLSVtotal", "male", "SUD", "PDB", "education", "MDFDR", "prevcrimeall")

# SPECT predictors 
spectList <- c("cerebellumdx", "cerebellumsin", "frontaldx", "frontalsin", "temporaldx", "temporalsin", "parietaldx", "parietalsin")

# pharmacological treatment variables
pharmaList <- c("antipsychotic", "antidepressant", "benzodiazepine", "nonbenzodiazepine", "anticholinergic")

# other variables
otherList <- c("timeatrisk", "primarydsmdiagnosis", "numberofaxis1", "numberofaxis2", "stillundercare", "totaldaysundercare2")

# primary DSM-IV diagnosis
primaryList <- c("primarydsm_psychosis", "primarydsm_mood", "primarydsm_pd", "primarydsm_DDA", "primarydsm_NDD")
```

## Missing data

Look for missing data in the list of predictors and note how many rows (if any) contain missing values. Save that in a text file so it can be reported in text.

```{r}
# look for NAs in the predictor variables
missing <- describe(data[c(baselineList, spectList, pharmaList, otherList, primaryList)])
missingNames <- row.names(missing)
tempMissing <- NULL

# loop through list of predictor variables and save to temporary vector
for (i in 1:length(missingNames)) {
  tempMissing <- append(tempMissing,
                        paste("There is/are",
                              50-missing[i, 2], 
                              "row/s missing in the", 
                              missingNames[i],
                              "variable"))
}

# capture output
results_missing <- capture.output(tempMissing)

# save output file to to .txt
write(results_missing, file = "results/results_missing.txt")

# remove from environment
remove(i, missing, missingNames, tempMissing, results_missing)
```

## Drop missing data

```{r}
data <- data[complete.cases(data[, baselineList]), ]
```

# Data exploration

## Reoffending

```{r}
# relevel the outcome variable
contrasts(data[[outcomeVariable]]) <- contr.treatment(levels(data[[outcomeVariable]]),
   base = which(levels(data[[outcomeVariable]]) == "yes"))

# how many reoffended?
results_reoffending <- capture.output(cat(
  "In total,",
  sum(data[outcomeVariable] == "yes"),
  "patients reoffended overall, corresponding to",
  as.numeric(round(prop.table(table(data[outcomeVariable]))[2], 2) * 100),
  "percent of the sample.\n\n",
  "In total,",
  sum(data$reoffendviolent == "yes"),
  "patients reoffended violently, corresponding to",
  as.numeric(round(prop.table(table(data$reoffendviolent))[2], 2) * 100),
  "percent of the sample."))

# save output file to to .txt
write(results_reoffending, file = "results/results_reoffending.txt")

# remove from environment
remove(results_reoffending)
```

## Dead and deported

It is possible that not all patients were followed for the entire follow-up period; some may have died, and some may have been deported. Results are printed, captured and saved to a text file.

```{r}
# capture and save output
results_deadanddeported <- capture.output(cat(
  "In total,",
  sum(data$dead == "yes"),
  "patients died during their time at risk, and they were followed for mean =",
  as.numeric(round(describe(data$daystodeath)[3], 0)),
  "days, with sd =",
  as.numeric(round(describe(data$daystodeath)[4], 0)),
  "ranging between",
  as.numeric(round(describe(data$daystodeath)[8], 0)),
  "and",
  as.numeric(round(describe(data$daystodeath)[9], 0)),
  "days.",
  "\n\nIn addition",
  sum(data$deported == "yes"),
  "patient was deported during the follow-up, and was followed for",
  as.numeric(describe(data$daystodeportation)[3]),
  "days."))

# write to .txt
write(results_deadanddeported, file = "results/results_deadanddeported.txt")

# remove from environment
remove(results_deadanddeported)
```

## Time at risk and age

```{r}
# capture and save output
results_timeatriskandage <- capture.output(cat(
  "In the entire sample, patients were followed for mean =",
  as.numeric(round(describe(data$timeatrisk)[3]), 0),
  "days, with sd =",
  as.numeric(round(describe(data$timeatrisk)[4]), 0),
  "ranging between",
  as.numeric(round(describe(data$timeatrisk)[8]), 0),
  "and",
  as.numeric(round(describe(data$timeatrisk)[9]), 0),
  "days.\n\nPatient's age at the time of FPI was median =",
  as.numeric(round(describe(data$age)[5], 2)),
  "years, with sd =",
  as.numeric(round(describe(data$age)[4], 2)),
  "ranging between",
  as.numeric(round(describe(data$age)[8], 2)),
  "and",
  as.numeric(round(describe(data$age)[9], 2))))

# write to .txt
write(results_timeatriskandage, file = "results/results_timeatriskandage.txt")

# remove from environment
remove(results_timeatriskandage)
```

## Diagnostic overview

```{r}
# capture and save output
results_diagnostics <- capture.output(cat(
  "In total,",
  sum(data$numberofaxis1 > 1),
  "patients (",
  round(sum(data$numberofaxis1 > 1) / nrow(data), 2)*100,
  "percent) had more than one DSM-IV Axis I disorder, with a range of",
  range(data$numberofaxis1)[1],"to",range(data$numberofaxis1)[2],
  "\n\nIn total,",
  sum(data$numberofaxis2 > 1),
  "patients (",
  round(sum(data$numberofaxis2 > 1) / nrow(data), 2)*100,
  "percent) had more than one DSM-IV Axis II disorder."))

# write to .txt
write(results_diagnostics, file = "results/results_diagnostics.txt")

# remove from environment
remove(results_diagnostics)
```

## Committed crimes

```{r}
# this is a very ugly solution, but it works
# I guess it could be done more elegantly with magrittr piping?
crimecategory <- sprintf("cat%d", 1:51)
crimecategory <- subset(data, select = crimecategory)
crimecategory <- data.frame(lapply(crimecategory, as.factor), stringsAsFactors=FALSE)
crimecategory <- as.matrix(crimecategory)
crimecategory <- as.vector(crimecategory)
crimecategory <- gsub(" ", "", crimecategory)
crimecategory[!nzchar(crimecategory)] <- NA
crimecategory <- crimecategory[!is.na(crimecategory)]
crimecategory <- crimecategory[!duplicated(crimecategory)]

# capture and save output
results_crimes <- capture.output(cat(
  "Crimes committed during the follow-up:\n",
  paste(crimecategory, collapse = "\n")))

# write to .txt
write(results_crimes, file = "results/results_crimes.txt")

# remove from environment
remove(results_crimes, crimecategory)
```

## Correlations

```{r}
source("scripts/cor.R")

# write to .txt
write.table(corDat, "results/results_correlations.txt", sep = "\t", row.names = FALSE)
remove(cList, corDat)
```

# Subset data

Subset the data to remove everything we don't need for subsequent analyses.

```{r}
# select only columns that match the variable names in our lists
idx <- match(c(outcomeVariable, baselineList, spectList, pharmaList, primaryList, otherList), names(data))

# subset
data <- data[, idx] 

remove(idx)
```

# Table 1

*Table 1* consists of an overview of all relevant variables. See *scripts/table_1.R* and *scripts/table_1_categorical.R* for details.

```{r message=TRUE, warning=TRUE, include=FALSE}
# source table scripts
source("scripts/table_1_categorical.R")
source("scripts/table_1.R")

# att row names as a column
table1$variable <- row.names(table1)

# reorder columns
table1 <- table1[, c(6, 1, 2, 3, 4, 5)]

# add a row indicating group, bind together
group <- c("", "All (n = 44)", "Non-recidivists (n = 28)", "Recidivists (n = 16)", "", "")
table1 <- rbind(group, table1)

write.table(table1, "results/table_1.txt", sep = "\t", row.names = FALSE)

# keep environment clean
remove(table1, group)
```

# Random Forest classification

## Predictors

Define which predictors to use in each step.

```{r}
# main analisys
baselinePredictors <- baselineList
extendedPredictors <- c(baselineList, spectList)

# supplementary analysis
pharmaPredictors <- c(baselineList, spectList, pharmaList)
```

## Formula specification

```{r}
formula_baseline <- paste(outcomeVariable, paste(baselinePredictors, collapse = "+"), sep = "~")
formula_extended <- paste(outcomeVariable, paste(extendedPredictors, collapse = "+"), sep = "~")
formula_pharma <- paste(outcomeVariable, paste(pharmaPredictors, collapse = "+"), sep = "~")
```

## Build models

Models are built using the [randomForest](https://cran.r-project.org/web/packages/randomForest/index.html) package with 10 000 trees each. Down-sampling is used so that each tree uses an equal number of offenders versus reoffenders. 

```{r}
# primary analysis
set.seed(seed)
baselineModel <- randomForest(as.formula(formula_baseline),
                      data = data,
                      sampsize = c(16, 16),
                      importance = TRUE,
                      ntree = 10000,
                      keep.inbag = TRUE,
                      proximity = TRUE,
                      keep.forest = TRUE)

set.seed(seed)
extendedModel <- randomForest(as.formula(formula_extended),
                      data = data,
                      sampsize = c(16, 16),
                      importance = TRUE,
                      ntree = 10000,
                      keep.inbag = TRUE,
                      proximity = TRUE,
                      keep.forest = TRUE)

# supplementary analysis
set.seed(seed)
pharmaModel <- randomForest(as.formula(formula_pharma),
                      data = data,
                      sampsize = c(16, 16),
                      importance = TRUE,
                      ntree = 10000,
                      keep.inbag = TRUE,
                      proximity = TRUE,
                      keep.forest = TRUE)

# keep environment clean
remove(formula_baseline, formula_extended, formula_pharma)
```

# Results

## Create confusion matrices

```{r}
# predict responses
baselinePred <- predict(baselineModel, type = "response")
extendedPred <- predict(extendedModel, type = "response")
pharmaPred <- predict(pharmaModel, type = "response")

# use 'caret' to create confusion matrix and associated metrics
baselineMatrix <- confusionMatrix(baselinePred, data$reoffendgeneral, positive = "yes")
extendedMatrix <- confusionMatrix(extendedPred, data$reoffendgeneral, positive = "yes")
pharmaMatrix <- confusionMatrix(pharmaPred, data$reoffendgeneral, positive = "yes")

# keep environment clean
remove(baselinePred, extendedPred, pharmaPred)
```

## Calculate AUCs

```{r}
# baseline model
baselinePredictions <- as.vector(baselineModel$votes[, 2])
baselinePredictions <- prediction(baselinePredictions, data$reoffendgeneral)
baselineAUC <- performance(baselinePredictions, "auc")
baselineAUC <- baselineAUC@y.values[[1]]

# extended model
extendedPredictions <- as.vector(extendedModel$votes[, 2])
extendedPredictions <- prediction(extendedPredictions, data$reoffendgeneral)
extendedAUC <- performance(extendedPredictions, "auc")
extendedAUC <- extendedAUC@y.values[[1]]

# pharma model
pharmaPredictions <- as.vector(pharmaModel$votes[, 2])
pharmaPredictions <- prediction(pharmaPredictions, data$reoffendgeneral)
pharmaAUC <- performance(pharmaPredictions, "auc")
pharmaAUC <- pharmaAUC@y.values[[1]]

# keep environment clean
remove(baselinePredictions, extendedPredictions, pharmaPredictions)
```

## Calculate Brier score

```{r}
baselinePred <- predict(baselineModel, type = "prob")[, 2]
extendedPred <- predict(extendedModel, type = "prob")[, 2]
pharmaPred <- predict(pharmaModel, type = "prob")[, 2]

resp <- as.numeric(data$reoffendgeneral) - 1

# scaled
baselineBrierScaled <- BrierScore(resp, baselinePred, scaled = TRUE)
extendedBrierScaled <- BrierScore(resp, extendedPred, scaled = TRUE)
pharmaBrierScaled <- BrierScore(resp, pharmaPred, scaled = TRUE)

remove(baselinePred, extendedPred, pharmaPred, resp)
```

## Model performance metrics

All relevant performance metrics are gathered and printed to a table.

```{r}
# store accuracy and related metrics
baselineAccuracy <- as.data.frame(baselineMatrix[[3]])
extendedAccuracy <- as.data.frame(extendedMatrix[[3]])
pharmaAccuracy <- as.data.frame(pharmaMatrix[[3]])

colnames(baselineAccuracy) <- c("Baseline")
colnames(extendedAccuracy) <- c("Extended")
colnames(pharmaAccuracy) <- c("Pharma")

# store additional metrics
baselineMetrics <- as.data.frame(baselineMatrix[[4]])
extendedMetrics <- as.data.frame(extendedMatrix[[4]])
pharmaMetrics <- as.data.frame(pharmaMatrix[[4]])

colnames(baselineMetrics) <- c("Baseline")
colnames(extendedMetrics) <- c("Extended")
colnames(pharmaMetrics) <- c("Pharma")

# bind together
results <- rbind(cbind(baselineAccuracy, extendedAccuracy, pharmaAccuracy),
                 cbind(baselineMetrics, extendedMetrics, pharmaMetrics))

# AUC and Brier scores added with a quick and dirty fix
AUCresults <- NULL
AUCresults$Baseline <- baselineAUC
AUCresults$Extended <- extendedAUC
AUCresults$Pharma <- pharmaAUC
AUCresults <- as.data.frame(AUCresults)

brierresultsscaled <- NULL
brierresultsscaled$Baseline <- baselineBrierScaled
brierresultsscaled$Extended <- extendedBrierScaled
brierresultsscaled$Pharma <- pharmaBrierScaled
brierresultsscaled <- as.data.frame(brierresultsscaled)

results <- rbind(brierresultsscaled, AUCresults, results)
row.names(results)[1] <- "Scaled Brier score"
row.names(results)[2] <- "AUC"
results$statistic <- row.names(results)
results <- results[, c(4, 1, 2, 3)]

# final column names fix
colnames(results) <- c("Statistic", "Baseline", "Extended", "Pharma")

# keep only the statistics we're reporting in text (optional, may wish to keep everything)
results <- results[c("Scaled Brier score", "AUC", "Accuracy", "AccuracyLower", "AccuracyUpper", "Sensitivity", "Specificity", "Pos Pred Value", "Neg Pred Value"), ]

# keep environment clean
remove(baselineAccuracy, extendedAccuracy, pharmaAccuracy,
       baselineMetrics, extendedMetrics, pharmaMetrics,
       baselineAUC, extendedAUC, pharmaAUC,
       baselineMatrix, extendedMatrix, pharmaMatrix,
       AUCresults)
```

## Save results

```{r}
write.table(results, "results/RF_results.txt", sep = "\t", row.names = FALSE)
remove(results)
```

# Figures

## Figure 2

Figure 2 in the article visualizes variable importance in both models. See *scripts/figurescript.R* for more details.

```{r}
# for quick cosmetic changes, set various text sizes here
textsize <- 2.5
axissize <- 7
axistitlesize <- 7
legendsize <- 7
legendtitlesize <- 7
titlesize <- 7

# source scripts
source("scripts/figurescript.R")

# combine the two plots into one
plots <- plot_grid(p1, p2, nrow = 1, ncol = 2, labels = c("Baseline model", "Extended model"), label_size = 9, hjust = -0.1, label_fontface = "plain")

# save in TIFF format
tiff("figures/figure2.tiff", width = 7, height = 2.8, units = "in", res = 400, compression = "none")
plots
dev.off()

# tidy up
remove(p1, p2, plots)
```

## Figure 3

Figure 3 in the manuscript is the partial dependence plots for the eight most important variables in the Baseline/Extended models. See *scripts/partialdepplot.R* for more details.

### Baseline model

```{r}
# select the top 8 predictors from Baseline model and manually add formatted names
partialPredictors <- NULL
partialPredictors$loop <- row.names(MDA1[order(MDA1$MeanDecreaseAccuracy, decreasing = TRUE)[1:8], ])
partialPredictors$names <- c("Cluster B personality disorder",
                             "Age at first crime",
                             "Age",
                             "Substance use disorder",
                             "PCL:SV total score",
                             "Previous criminality",
                             "Educational attainment",
                             "Male sex")

# specify baseline model
model <- baselineModel

# source pdp plot script
source("scripts/partialdepplot.R")

# arrange in grid
pdpbaseline <- plot_grid(plot[[1]], plot[[2]], plot[[3]], plot[[4]],
                   plot[[5]], plot[[6]], plot[[7]], plot[[8]],
                   ncol = 4, nrow = 2)
```

### Extended model

```{r}
# select the top 8 predictors from Extended model and manually add formatted names
partialPredictors <- NULL
partialPredictors$loop <- row.names(MDA2[order(MDA2$MeanDecreaseAccuracy, decreasing = TRUE)[1:8], ])

partialPredictors$names <- c("Right parietal lobe rCBF",
                             "Cluster B personality disorder",
                             "Age at first crime",
                             "Left parietal lobe rCBF",
                             "Age",
                             "Left temporal lobe rCBF",
                             "Right cerebellum rCBF",
                             "Right temporal lobe rCBF")

# specify extended model
model <- extendedModel

# source pdp plot script
source("scripts/partialdepplot.R")

# arrange in grid
pdpextended <- plot_grid(plot[[1]], plot[[2]], plot[[3]], plot[[4]],
                   plot[[5]], plot[[6]], plot[[7]], plot[[8]],
                   ncol = 4, nrow = 2)
```

### Combined figure

```{r}
# combine baseline and extended plots
pdpcombined <- plot_grid(pdpbaseline, pdpextended, nrow = 2, ncol = 1, labels = c("Baseline model", "Extended model"), label_size = 9, hjust = -0.1, label_fontface = "plain")

# save in TIFF format
tiff("figures/figure3.tiff", width = 6, height = 5.5, units = "in", res = 400, compression = "none")
pdpcombined
dev.off()

# keep environment clean
remove(MDA1, MDA2, numeric, categorical, i, var, plot, pdpbaseline, pdpextended, pdpcombined, temp, axissize, axistitlesize, legendsize, legendtitlesize, textsize, titlesize, model)
```

# Session info

Write a text file with the session info, for future reference.

```{r}
sessInfo <- capture.output(sessionInfo())
write(sessInfo, file = "sessioninfo.txt")

remove(sessInfo)
```

# Stop timer

Total running time is included for fun.

```{r}
runningtime <- capture.output(proc.time() - timer)
write(runningtime, file = "runningtime.txt")
```
